---
title: "Group_8_Analysis"
format: html
editor: visual
---

```{r}
#| echo: false
#| warning: false
#| message: false
library(ggplot2)
library(tidyverse)
library(gt)
library(broom)
library(knitr)
library(dplyr)
library(tibble)
```


# Which film properties influence whether a film receives an IMDB rating greater than 7? {#sec-study}


# 1 Research Question

The aim of this analysis is to examine which film characteristics are associated with whether a film receives an IMDb rating greater than 7 or not.

To address this question, a binary response variable is considered, taking the value 1 if a film has an IMDb rating above 7 and 0 otherwise. The analysis focuses on whether release year, film length, production budget, number of votes, and genre are related to the probability of receiving a high rating.

Given that the response variable is binary, the analysis employs logistic regression as part of the Generalised Linear Model framework.





# 2 The Dataset {#sec-dataset}

The dataset used in this analysis comes from the IMDB film database, which provides a variety of information on films that have been released. The data used here represent a subset of the original database.

## 2.1 Inspect the dataset {#sec-insp}

```{r}
#| echo: false
#| warning: false
#| message: false

# Load the dataset from the CSV file
movies <- read.csv("dataset08.csv")
```

After loading the dataset, the first few observations are displayed to provide an initial overview of the data.

```{r}
#| echo: false
#| warning: false
#| message: false

#Display the first six rows of the dataset
head(movies)
```

The structure of the dataset is examined to identify the variables and their data types.

```{r}
#| echo: false
#| warning: false
#| message: false
#| results: false

#Check the structure of the dataset
str(movies)
```

Based on the structure output, the dataset contains 2,847 observations and seven variables. The variables include a mixture of integer, numeric, and character types. Specifically, film_id, year, length, and votes are stored as integer variables, while budget and rating are numeric variables. The genre variable is recorded as a character variable representing the film category. It also suggests that some variables (e.g., length) contain missing values (NA).

To further explore the distribution of the variables in the dataset, summary statistics are computed.

```{r}
#| echo: false
#| warning: false
#| message: false
#| results: false

# Summary statistics of variables
summary(movies)
```


```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-numvars
#| tbl-cap: "Summary statistics for numerical variables"

vars <- movies %>% select(film_id, year, length, budget, votes, rating)

numeric_stats <- rbind(
  summarise(vars, across(everything(), ~min(.,            na.rm = TRUE))),
  summarise(vars, across(everything(), ~quantile(., 0.25, na.rm = TRUE))),
  summarise(vars, across(everything(), ~median(.,         na.rm = TRUE))),
  summarise(vars, across(everything(), ~mean(.,           na.rm = TRUE))),
  summarise(vars, across(everything(), ~quantile(., 0.75, na.rm = TRUE))),
  summarise(vars, across(everything(), ~max(.,            na.rm = TRUE))),
  summarise(vars, across(everything(), ~sum(is.na(.))))
) %>%
  t() %>%
  as.data.frame() %>%
  setNames(c("Min", "1st Qu", "Median", "Mean", "3rd Qu", "Max", "Missing")) %>%
  rownames_to_column("Variable") %>%
  mutate(across(-Variable, ~round(as.numeric(.), 2)))

kable(numeric_stats, caption = "Summary statistics of numerical variables")
```

Table @tbl-numvars presents the summary statistics for the numerical variables included in the analysis.

As shown in Table @tbl-numvars, the release year of the films spans from 1898 to 2005, suggesting that the dataset covers a long historical period. Film length varies ranging from 1 to 480 minutes. The production budget ranges between 2.5 and 22.3, while the number of votes shows particularly large variation, with a maximum value of 149,494. In contrast, film ratings appear to be relatively concentrated, with values ranging from 0.8 to 9.2. It is also noteworthy that the variable *length* contains 131 missing observations, whereas the remaining numerical variables do not exhibit missing values.

In addition to the numerical variables, the distribution of film genres is also examined. 

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-genre
#| tbl-cap: "Distribution of films across genres"

genre_summary <- movies %>%
  count(genre, sort = TRUE)

kable(genre_summary, caption = "Number of films by genre")
```

Table @tbl-genre reports the number of films in each genre.

Table @tbl-genre presents the distribution of films across seven genres in the dataset. Action, Drama, and Comedy account for the largest number of observations, whereas genres such as Animation, Documentary, Short, and Romance appear less frequently. This suggests that the dataset is dominated by a few major genres.

## 2.2 Variable descriptions

In this dataset, each film is represented by several variables describing its main characteristics. Table @tbl-var-desc shows the main variables used, including their data types and brief descriptions.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-variable
#| tbl-cap: "Description of variables in the dataset"
#| 
var_desc <- data.frame(
  Variable = c("film_id", "year", "length", "budget", "votes", "genre", "rating"),
  Type = c("Integer", "Integer", "Numeric", "Numeric", "Integer", "Character", "Numeric"),
  Description = c(
    "The unique identifier for the film",
    "Year of release of the film in cinemas",
    "Duration (in minutes)",
    "Budget for the films production (in $1000000s)",
    "Number of positive votes received by viewers",
    "Genre of the film",
    "IMDB rating from 0-10"
  )
)

kable(var_desc)
```

Table @tbl-variable shows the variables included in the dataset.

The dataset contains both numerical and categorical variables describing different aspects of each film. These variables will later be used as predictors in the statistical model.



# 3 Exploratory Analysis

Before fitting the GLM, exploratory analysis was carried out to examine the relationships between the response variable and the candidate predictors. This helps to identify broad patterns in the data and provides preliminary support for variable selection in the modelling stage.


```{r}
#| echo: false
#| warning: false
#| message: false
# Create binary response variable
movies$high_rating <- ifelse(movies$rating > 7, 1, 0)
```


### 3.1 Release Year and High IMDB Ratings

Release year may influence how films are rated. To explore this, films were grouped by release decade and the proportion of highly rated films (IMDB rating \> 7) was calculated for each group.

Very early decades contained only a small number of films, so films released before 1910 were excluded from this plot to provide a more stable comparison across release periods.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-explo-year

movies_year <- movies %>%
  filter(year >= 1910) %>%
  mutate(decade = floor(year/10)*10) %>%
  group_by(decade) %>%
  summarise(
    proportion_high = mean(high_rating, na.rm = TRUE),
    count = n()
  )

ggplot(movies_year, aes(x = decade, y = proportion_high)) +
  geom_line(colour = "mediumpurple", linewidth = 1.2) +
  geom_point(colour = "gold2", size = 2.5) +
  geom_hline(yintercept = mean(movies$high_rating),
             linetype = "dashed",
             colour = "coral",
             linewidth = 0.8) +
  labs(
    title = "Trend in the Proportion of Highly Rated Films by Decade",
    x = "Decade",
    y = "Proportion of Films with Rating > 7"
  ) +
  theme_bw()
```

Figure @fig-explo-year shows the proportion of highly rated films across release decades. The dashed horizontal line represents the overall proportion of films with ratings above 7 in the dataset.

The proportion of highly rated films appears to vary across decades. Earlier decades tend to show somewhat higher proportions, while films released between the 1950s and 1980s appear to have relatively lower proportions. However, some early decades contain relatively few films, so these proportions should be interpreted with caution.



### 3.2 Film Length and IMDB Ratings

Film length may influence audience perception of a movie. To explore this relationship, film length is grouped into five categories using quantile-based cut-offs so that each group contains approximately 20% of the films in the dataset.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-explo-length

movies_length <- subset(movies, !is.na(length) & !is.na(rating) & !is.na(high_rating))

# Create film length groups
length_breaks <- quantile(movies_length$length, probs = seq(0, 1, 0.2), na.rm = TRUE)

movies_length$length_group <- cut(
  movies_length$length,
  breaks = length_breaks,
  include.lowest = TRUE,
  labels = c("Very Short", "Short", "Medium", "Long", "Very Long")
)

# Calculate proportion of high ratings in each length group
length_prop <- aggregate(
  high_rating ~ length_group,
  data = movies_length,
  FUN = mean
)

ggplot(movies_length, aes(x = length, y = high_rating)) +
  geom_jitter(height = 0.05, alpha = 0.3, colour = "pink") +
  geom_smooth(method = "glm",
              method.args = list(family = "binomial"),
              colour = "navy") +
  labs(
    title = "Relationship Between Film Length and Probability of High Rating",
    x = "Film Length (minutes)",
    y = "Probability of High Rating"
  ) +
  theme_bw()
```

Figure @fig-explo-length shows the proportion of highly rated films across film length groups. The overall proportion of films with ratings above 7 is shown by the dashed horizontal line.

Very short films appear to have a higher proportion of ratings above 7 than the other length groups, while the remaining groups are more similar to one another. This suggests that film length may be associated with high ratings, although the pattern may partly reflect differences in genre composition across length groups.



### 3.3 Production budget and IMDB Ratings

Film budget may also influence how audiences perceive a movie. To explore this possibility, the distribution of film budgets and their relationship with IMDB ratings were examined.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-explo-budget1
#| fig-cap: Relationship between film budget and the probability of receiving a high IMDb rating

ggplot(movies,
       aes(x = budget,
           y = high_rating)) +
  geom_jitter(height = 0.05, alpha = 0.3) +
  geom_smooth(
    method = "glm",
    method.args = list(family="binomial"),
    colour = "red"
  ) +
  theme_bw()

```

Figure @fig-explo-budget shows the relationship between film production budget and the probability of receiving a high IMDb rating. Each point represents an individual film, and the red curve corresponds to the fitted logistic regression line.

The fitted curve shows a gradual upward trend, suggesting that films with larger budgets may be somewhat more likely to receive high ratings. This suggests that budget may be positively associated with the likelihood of a film receiving a high rating.


### 3.4 Votes and High IMDB Ratings

The number of votes may reflect a film’s popularity or level of audience engagement. To explore this relationship, films were grouped based on their number of votes and the proportion of highly rated films (IMDB rating \> 7) was calculated for each group.

The approximate vote ranges for the five groups are:\
Very Low: 5-10 votes\
Low: 10-20 votes\
Medium: 20-47 votes\
High: 47-174 votes\
Very High: above 174 votes.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-explo-votes
#| fig-cap: Proportion of High IMDB Ratings Across Vote Groups

# Calculate vote quantiles
vote_breaks <- quantile(movies$votes, probs = seq(0,1,0.2), na.rm = TRUE)

# Create vote groups with readable labels
movies$vote_group <- cut(
  movies$votes,
  breaks = vote_breaks,
  include.lowest = TRUE,
  labels = c("Very Low", "Low", "Medium", "High", "Very High")
)

# Calculate proportion of high ratings in each vote group
prop_table <- aggregate(high_rating ~ vote_group,
                        data = movies,
                        FUN = mean)

ggplot(prop_table, aes(x = vote_group, y = high_rating)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_hline(yintercept = mean(movies$high_rating),
             colour = "red",
             linetype = "dashed",
             linewidth = 1) +
  labs(
    title = "Proportion of High IMDB Ratings Across Vote Groups",
    x = "Vote Groups",
    y = "Proportion of Films with Rating > 7"
  ) +
  theme_minimal()
```


Figure @fig-explo-votes shows the proportion of highly rated films across vote groups. The dashed horizontal line represents the overall proportion of films with ratings above 7 in the dataset.

The proportion of highly rated films appears to vary across vote groups, with lower-vote groups showing slightly higher proportions than higher-vote groups. Although the pattern is not especially strong, it suggests that the relationship between vote count and high ratings may not be straightforward.



### 3.5 Film Genre and High IMDB Ratings

Film genre may influence how movies are perceived by audiences. Different genres may attract different audiences and may vary in storytelling style, which could affect IMDB ratings.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-explo-genre

# Calculate proportion of highly rated films for each genre
genre_prop <- movies %>%
  group_by(genre) %>%
  summarise(
    proportion_high = mean(high_rating, na.rm = TRUE),
    count = n()
  )

ggplot(genre_prop,
       aes(x = reorder(genre, proportion_high),
           y = proportion_high)) +
  geom_col(fill = "seagreen") +
  geom_hline(yintercept = mean(movies$high_rating, na.rm = TRUE),
             colour = "darkred",
             linetype = "dashed",
             linewidth = 1) +
  coord_flip() +
  labs(
    title = "Proportion of Highly Rated Films by Genre",
    x = "Genre",
    y = "Proportion of Films with Rating > 7"
  ) +
  theme_bw()
```


Figure @fig-explo-genre shows substantial variation in the proportion of highly rated films across genres. Some genres have much higher proportions of films with ratings above 7 than others.

This suggests that genre may be an important predictor of high ratings. However, some genres contain relatively few observations, so the corresponding proportions may be less stable and should be interpreted cautiously.





## Selection of Candidate Variables

Based on the exploratory analysis, several variables were identified as potential predictors of whether a film receives an IMDb rating greater than 7. These variables include:\
- votes\
- film length\
- release year\
- genre.

These variables will be used as candidate predictors in the logistic regression model.

## Model Fitting

To investigate the factors associated with whether a film receives a high IMDB rating, a logistic regression model was fitted.

The response variable is a binary indicator of whether a film has a rating greater than 7. The candidate predictor variables identified in the exploratory analysis include the number of votes, film length, release year, and genre.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-glm-full

# Fit logistic regression model
model_full <- glm(
  high_rating ~ votes + length + year + genre,
  data = movies,
  family = binomial
)

tidy(model_full, exponentiate = TRUE) %>%
  kable(
    digits = 3,
    col.names = c("Variable","Odds Ratio","Std Error","z value","p value"),
    caption = "Logistic regression results (full model)"
  )
```

The logistic regression results are summarized in @tbl-glm-full. Film length, release year, and several genres show statistically significant associations with the probability of receiving a high IMDB rating (p \< 0.05).

However, the number of votes is not statistically significant (p = 0.949). Therefore, the variable votes was removed and a simplified model was fitted.

### Simplified Logistic Regression Model

Since the number of votes was not statistically significant in the full model, it was removed and a simplified logistic regression model was fitted using the remaining predictors: film length, release year, and genre.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-glm-final

# Fit simplified logistic regression model
model_final <- glm(
  high_rating ~ length + year + genre,
  data = movies,
  family = binomial
)

tidy(model_final) %>%
  kable(
    digits = 3,
    col.names = c("Variable","Estimate","Std Error","z value","p value"),
    caption = "Logistic regression results (simplified model)"
  )
```

@tbl-glm-final indicates that film length and release year are significant predictors of whether a film receives a high IMDb rating (p \< 0.001).

Longer films tend to be less likely to receive ratings above 7, while more recently released films are slightly more likely to receive higher ratings.

In addition, differences across genres suggest that the likelihood of receiving high ratings varies depending on film type.

### Model Comparison

To better illustrate the effect of variable selection, the coefficients of the full model and the simplified model are compared in @glm-sum.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-glm-sum

full <- tidy(model_full) %>%
  select(term, estimate) %>%
  rename(Full_Model = estimate)
final <- tidy(model_final) %>%
  select(term, estimate) %>%
  rename(Final_Model = estimate)

comparison <- full |> 
  full_join(final, by = "term")

kable(comparison, digits = 3,
      col.names = c("Predictor", "Full Model", "Final Model"),
      caption = "Comparison of coefficients between full and simplified models")
```

@tbl-glm-sum compares the estimated coefficients between the full model and the simplified model.

The results show that removing the non-significant variable (votes) does not substantially change the estimated coefficients of the remaining predictors. This suggests that the simplified model provides a more parsimonious representation while maintaining similar estimated effects.

## Predicted Probabilities

To further illustrate the fitted logistic regression model, predicted probabilities were calculated across a range of film lengths while holding release year constant at its mean value.

Predicted probabilities were plotted for several selected genres to visualize how the probability of receiving a high IMDB rating varies with film length and genre.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-pred

# Create new data for prediction
newdata <- expand.grid(
  length = seq(0, 200, length.out = 100),
  year = mean(movies$year, na.rm = TRUE),
  genre = c("Comedy", "Documentary", "Drama", "Animation")
)

# Compute predicted probabilities from the simplified model
newdata$pred_prob <- predict(model_final, newdata = newdata, type = "response")

# Plot predicted probabilities
ggplot(newdata, aes(x = length, y = pred_prob, colour = genre)) +
  geom_line(linewidth = 1.2) +
  labs(
    title = "Predicted Probability of High IMDB Rating by Film Length and Genre",
    x = "Film Length (minutes)",
    y = "Predicted Probability",
    colour = "Genre"
  ) +
  theme_bw()
```

@tbl-pred shows the predicted probability of receiving a high IMDb rating across different film lengths for selected genres.

Release year is not shown in the figure because it was held constant at its mean value when calculating predicted probabilities. This allows the effects of film length and genre on the predicted probability to be visualized more clearly while still using the full simplified model.

The predicted probability generally decreases as film length increases, indicating that longer films are less likely to receive ratings above 7. Differences across genres are also evident, with some genres consistently showing higher predicted probabilities than others across the range of film lengths.

## Interpretation of Odds Ratios

To aid interpretation of the logistic regression model, the estimated coefficients were transformed into odds ratios by taking their exponential values.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-odds_ratio

tidy(model_final) %>%
  mutate(odds_ratio = exp(estimate)) %>%
  select(term, estimate, odds_ratio, p.value) %>%
  kable(
    digits = 3,
    col.names = c("Predictor", "Estimate", "Odds Ratio", "p value"),
    caption = "Odds ratios from the simplified logistic regression model"
  )
```

@tbl-odds_ratio presents the odds ratios from the simplified logistic regression model.

Film length shows an odds ratio below 1, indicating that longer films are less likely to receive ratings above 7. In contrast, more recent films and some genres have higher odds of receiving high ratings.

## Model Limitation

Although the logistic regression model provides useful insights, several limitations should be noted.

First, the analysis is based on observational data and therefore cannot establish causal relationships between predictors and IMDb ratings.

Second, some genres contain relatively small numbers of films, which may affect the stability of the estimated coefficients.

Finally, other potentially important factors, such as director reputation or production budget, were not included in the dataset and may also influence film ratings.

## Conclusion

This study examined factors associated with whether a film receives a high IMDB rating.

The results of the logistic regression analysis suggest that film length, release year, and genre are important predictors of high ratings. Longer films tend to have lower probabilities of receiving ratings above 7, while more recently released films show slightly higher probabilities. In addition, some genres such as comedy and documentary are more likely to receive higher ratings compared with the reference genre.

Overall, the findings suggest that both film characteristics and genre differences play a role in determining audience ratings.
