---
title: "Group_8_Analysis"
format:
  html:
    theme: cosmo
    toc: true
    number-sections: true
editor: visual
---

```{r}
#| echo: false
#| warning: false
#| message: false

# Load required packages for data manipulation, visualisation and modelling
library(ggplot2)     # plotting
library(tidyverse)   # data manipulation tools
library(gt)          # table formatting
library(broom)       # converting model outputs into tidy data frames
library(knitr)       # table display in reports
library(dplyr)       # data wrangling functions
library(tibble)      # tidy data frame utilities
```

# Introduction

This study investigates which film properties influence whether a film receives an IMDb rating greater than 7. IMDb ratings are widely used as an indicator of audience perception and overall film quality, making them an interesting outcome to analyse.

Understanding which characteristics are associated with higher ratings may provide insights into what factors contribute to the perceived success of a film. In particular, features such as release year, film length, number of votes and genre may influence how audiences evaluate movies.

To examine this question, a binary response variable is constructed indicating whether a film has an IMDb rating above 7. A logistic regression model within the Generalised Linear Model framework is then used to analyse the relationship between film characteristics and the probability of receiving a high rating.

# Research Question

The aim of this analysis is to examine which film characteristics are associated with whether a film receives an IMDb rating greater than 7 or not.

To address this question, a binary response variable is considered, taking the value 1 if a film has an IMDb rating above 7 and 0 otherwise. The analysis focuses on whether release year, film length, number of votes, and genre are related to the probability of receiving a high rating.

# The Dataset {#sec-dataset}

The dataset used in this analysis comes from the IMDB film database, which provides a variety of information on films that have been released. The data used here represent a subset of the original database.

## Inspect the dataset {#sec-insp}

```{r}
#| echo: false
#| warning: false
#| message: false

# Load the dataset from the CSV file
movies <- read.csv("dataset08.csv")
```

After loading the dataset, the first few observations are displayed to provide an initial overview of the data.

```{r}
#| echo: false
#| warning: false
#| message: false

# Display the first six rows of the dataset
# This helps us quickly inspect the structure and variables
head(movies)
```

The structure of the dataset is examined to identify the variables and their data types.

```{r}
#| echo: false
#| warning: false
#| message: false
#| results: false

# Check the structure of the dataset
# Shows variable types and basic information about each column
str(movies)
```

Based on the structure output, the dataset contains 2,847 observations and seven variables. The variables include a mixture of integer, numeric, and character types. Specifically, film_id, year, length, and votes are stored as integer variables, while budget and rating are numeric variables. The genre variable is recorded as a character variable representing the film category. It also suggests that some variables (e.g., length) contain missing values (NA).

To further explore the distribution of the variables in the dataset, summary statistics are computed.

```{r}
#| echo: false
#| warning: false
#| message: false
#| results: false

# Summary statistics of variables
# Provides basic descriptive statistics for each variable
summary(movies)
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-numvars
#| tbl-cap: "Summary statistics for numerical variables"

# Select numerical variables for summary statistics
vars <- movies %>% select(film_id, year, length, budget, votes, rating)

# Calculate common descriptive statistics
numeric_stats <- rbind(
  summarise(vars, across(everything(), ~min(.,            na.rm = TRUE))),   # Minimum
  summarise(vars, across(everything(), ~quantile(., 0.25, na.rm = TRUE))),   # First quartile
  summarise(vars, across(everything(), ~median(.,         na.rm = TRUE))),   # Median
  summarise(vars, across(everything(), ~mean(.,           na.rm = TRUE))),   # Mean
  summarise(vars, across(everything(), ~quantile(., 0.75, na.rm = TRUE))),   # Third quartile
  summarise(vars, across(everything(), ~max(.,            na.rm = TRUE))),   # Maximum
  summarise(vars, across(everything(), ~sum(is.na(.))))                      # Number of missing values
) %>%
  t() %>%                     # transpose table for readability
  as.data.frame() %>%
  setNames(c("Min", "1st Qu", "Median", "Mean", "3rd Qu", "Max", "Missing")) %>%
  rownames_to_column("Variable") %>%        # add variable names as column
  mutate(across(-Variable, ~round(as.numeric(.), 2)))  # round values

# Display the table
kable(numeric_stats)
```

@tbl-numvars presents the summary statistics for the numerical variables included in the analysis.

As shown in @tbl-numvars, the release year of the films spans from 1898 to 2005, suggesting that the dataset covers a long historical period. Film length varies ranging from 1 to 480 minutes. The production budget ranges between 2.5 and 22.3, while the number of votes shows particularly large variation, with a maximum value of 149,494. In contrast, film ratings appear to be relatively concentrated, with values ranging from 0.8 to 9.2. It is also noteworthy that the variable *length* contains 131 missing observations, whereas the remaining numerical variables do not exhibit missing values.

In addition to the numerical variables, the distribution of film genres is also examined.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-genre
#| tbl-cap: "Distribution of films across genres"

# Count number of films in each genre
genre_summary <- movies %>%
  count(genre, sort = TRUE)

# Display genre frequency table
kable(genre_summary, caption = "Number of films by genre")
```

@tbl-genre reports the number of films in each genre.

@tbl-genre presents the distribution of films across seven genres in the dataset. Action, Drama, and Comedy account for the largest number of observations, whereas genres such as Animation, Documentary, Short, and Romance appear less frequently. This suggests that the dataset is dominated by a few major genres.

## Variable descriptions

In this dataset, each film is represented by several variables describing its main characteristics. @tbl-variable shows the main variables used, including their data types and brief descriptions.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-variable
#| tbl-cap: "Description of variables in the dataset"
#| 
var_desc <- data.frame(
  Variable = c("film_id", "year", "length", "budget", "votes", "genre", "rating"),
  Type = c("Integer", "Integer", "Numeric", "Numeric", "Integer", "Character", "Numeric"),
  Description = c(
    "The unique identifier for the film",
    "Year of release of the film in cinemas",
    "Duration (in minutes)",
    "Budget for the films production (in $1000000s)",
    "Number of positive votes received by viewers",
    "Genre of the film",
    "IMDB rating from 0-10"
  )
)

kable(var_desc)
```

@tbl-variable shows the variables included in the dataset.

The dataset contains both numerical and categorical variables describing different aspects of each film. These variables will later be used as predictors in the statistical model.

# Exploratory Analysis

Before fitting the GLM, exploratory analysis was carried out to examine the relationships between the response variable and the candidate predictors. This helps to identify broad patterns in the data and provides preliminary support for variable selection in the modelling stage.

```{r}
#| echo: false
#| warning: false
#| message: false
# Create a binary response variable
# 1 = rating above 7
# 0 = rating 7 or below
movies$high_rating <- ifelse(movies$rating > 7, 1, 0)
```

### Release Year and High IMDB Ratings

Release year may influence how films are rated. To explore this, films were grouped by release decade and the proportion of highly rated films (IMDB rating \> 7) was calculated for each group.

Very early decades contained only a small number of films, so films released before 1910 were excluded from this plot to provide a more stable comparison across release periods.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-explo-year
#| fig-cap: "Proportion of Highly Rated Films by Release Decade"
#| fig-align: center
#| fig-width: 6
#| fig-height: 4
#| out-width: 75%

# Group films by decade and calculate proportion of high ratings
movies_year <- movies %>%
  filter(year >= 1910) %>%                # remove very early years with few observations
  mutate(decade = floor(year/10)*10) %>%  # create decade variable
  group_by(decade) %>%
  summarise(
    proportion_high = mean(high_rating, na.rm = TRUE), # proportion of rating > 7
    count = n()                                        # number of films
  )

# Plot proportion of high ratings by decade
ggplot(movies_year, aes(x = decade, y = proportion_high)) +
  geom_line(colour = "mediumpurple", linewidth = 1.2) +
  geom_point(colour = "gold2", size = 2.5) +
  geom_hline(yintercept = mean(movies$high_rating),    # overall average
             linetype = "dashed",
             colour = "coral",
             linewidth = 0.8) +
  labs(
    x = "Decade",
    y = "Proportion of Films with Rating > 7"
  ) +
  theme_bw()
```

@fig-explo-year shows the proportion of highly rated films across release decades. The dashed horizontal line represents the overall proportion of films with ratings above 7 in the dataset.

The proportion of highly rated films appears to vary across decades. Earlier decades tend to show somewhat higher proportions, while films released between the 1950s and 1980s appear to have relatively lower proportions. However, some early decades contain relatively few films, so these proportions should be interpreted with caution.

### Film Length and IMDB Ratings

Film length may influence audience perception of a movie. To explore this relationship, film length is treated as a continuous variable and the relationship between film length and the probability of receiving a high rating is visualised using a logistic smoothing curve.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-explo-length
#| fig-cap: "Relationship Between Film Length and the Probability of High IMDb Ratings"
#| fig-align: center
#| fig-width: 6
#| fig-height: 4
#| out-width: 75%

# Remove missing values before analysis
movies_length <- subset(movies, !is.na(length) & !is.na(rating) & !is.na(high_rating))

# Create film length groups
length_breaks <- quantile(movies_length$length, probs = seq(0, 1, 0.2), na.rm = TRUE)

movies_length$length_group <- cut(
  movies_length$length,
  breaks = length_breaks,
  include.lowest = TRUE,
  labels = c("Very Short", "Short", "Medium", "Long", "Very Long")
)

# Calculate proportion of high ratings in each length group
length_prop <- aggregate(
  high_rating ~ length_group,
  data = movies_length,
  FUN = mean
)

# Plot relationship between film length and probability of high rating
ggplot(movies_length, aes(x = length, y = high_rating)) +
  geom_jitter(height = 0.05, alpha = 0.3, colour = "pink") +  # scatter points
  geom_smooth(method = "glm",
              method.args = list(family = "binomial"),        # logistic smoothing curve
              colour = "navy") +
  labs(
    x = "Film Length (minutes)",
    y = "Probability of High Rating"
  ) +
  theme_bw()
```

@fig-explo-length shows the relationship between film length and the probability of receiving a high IMDb rating.

The probability of receiving a rating above 7 appears to be higher for shorter films and gradually decreases as film length increases. This suggests that film length may be associated with high ratings, although the pattern may partly reflect differences in genre composition across films of different lengths.

### Production budget and IMDB Ratings

Film budget may also influence how audiences perceive a movie. To explore this possibility, the distribution of film budgets and their relationship with IMDB ratings were examined.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-explo-budget1
#| fig-cap: Relationship between film budget and the probability of receiving a high IMDb rating
#| fig-align: center
#| fig-width: 6
#| fig-height: 4
#| out-width: 75%

# Visualise relationship between production budget and high ratings
ggplot(movies,
       aes(x = budget,
           y = high_rating)) +
  geom_jitter(height = 0.05, alpha = 0.3) +        # scatter points
  geom_smooth(
    method = "glm",
    method.args = list(family="binomial"),         # logistic regression curve
    colour = "red"
  ) +
  theme_bw()

```

@fig-explo-budget1 shows the relationship between film production budget and the probability of receiving a high IMDb rating. Each point represents an individual film, and the red curve corresponds to the fitted logistic regression line.

The fitted curve shows a gradual upward trend, suggesting that films with larger budgets may be somewhat more likely to receive high ratings. This suggests that budget may be positively associated with the likelihood of a film receiving a high rating.

### Votes and High IMDB Ratings

The number of votes may reflect a film’s popularity or level of audience engagement. Since the distribution of votes is highly skewed, with some films receiving far more votes than others, it is not easy to compare patterns directly using the raw vote counts. Therefore, films were divided into five groups based on vote quantiles, so that each group contains approximately the same number of observations. This makes it easier to compare the proportion of highly rated films across different levels of audience engagement.

The approximate vote ranges for the five groups are:\
Very Low: 5-10 votes\
Low: 10-20 votes\
Medium: 20-47 votes\
High: 47-174 votes\
Very High: above 174 votes.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-explo-votes
#| fig-cap: Proportion of High IMDB Ratings Across Vote Groups
#| fig-align: center
#| fig-width: 6
#| fig-height: 4
#| out-width: 75%

# Calculate vote quantiles to create groups
vote_breaks <- quantile(movies$votes, probs = seq(0,1,0.2), na.rm = TRUE)

# Create vote group variable
movies$vote_group <- cut(
  movies$votes,
  breaks = vote_breaks,
  include.lowest = TRUE,
  labels = c("Very Low", "Low", "Medium", "High", "Very High")
)

# Calculate proportion of high ratings in each group
prop_table <- aggregate(high_rating ~ vote_group,
                        data = movies,
                        FUN = mean)

# Plot proportions
ggplot(prop_table, aes(x = vote_group, y = high_rating)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_hline(yintercept = mean(movies$high_rating),
             colour = "red",
             linetype = "dashed",
             linewidth = 1) +
  labs(
    x = "Vote Groups",
    y = "Proportion of Films with Rating > 7"
  ) +
  theme_minimal()
```

@fig-explo-votes shows the proportion of highly rated films across vote groups. The dashed horizontal line represents the overall proportion of films with ratings above 7 in the dataset.

The proportion of highly rated films appears to vary across vote groups, with lower-vote groups showing slightly higher proportions than higher-vote groups. Although the pattern is not especially strong, it suggests that the relationship between vote count and high ratings may not be straightforward.

### Film Genre and High IMDB Ratings

Film genre may influence how movies are perceived by audiences. Different genres may attract different audiences and may vary in storytelling style, which could affect IMDB ratings.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-explo-genre
#| fig-cap: "Proportion of Highly Rated Films by Genre"
#| fig-align: center
#| fig-width: 6
#| fig-height: 4
#| out-width: 75%

# Calculate proportion of high ratings for each genre
genre_prop <- movies %>%
  group_by(genre) %>%
  summarise(
    proportion_high = mean(high_rating, na.rm = TRUE),
    count = n()
  )

# Plot genre comparison
ggplot(genre_prop,
       aes(x = reorder(genre, proportion_high),
           y = proportion_high)) +
  geom_col(fill = "seagreen") +
  geom_hline(yintercept = mean(movies$high_rating, na.rm = TRUE),
             colour = "darkred",
             linetype = "dashed",
             linewidth = 1) +
  coord_flip() +
  labs(
    x = "Genre",
    y = "Proportion of Films with Rating > 7"
  ) +
  theme_bw()
```

@fig-explo-genre shows substantial variation in the proportion of highly rated films across genres. Some genres have much higher proportions of films with ratings above 7 than others.

This suggests that genre may be an important predictor of high ratings. However, some genres contain relatively few observations, so the corresponding proportions may be less stable and should be interpreted cautiously.

# Logistic Regression Model
## Selection of Candidate Variables

Based on the exploratory analysis, several variables were identified as potential predictors of whether a film receives an IMDb rating greater than 7. These variables include:\
- votes\
- film length\
- release year\
- genre.

These variables will be used as candidate predictors in the logistic regression model.

## Model Fitting

To investigate the factors associated with whether a film receives a high IMDB rating, a logistic regression model was fitted.

The response variable is a binary indicator of whether a film has a rating greater than 7. The candidate predictor variables identified in the exploratory analysis include the number of votes, film length, release year, and genre.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-glm-full

# Fit logistic regression model including all candidate predictors
model_full <- glm(
  high_rating ~ votes + length + year + genre,
  data = movies,
  family = binomial
)

# Display model results with odds ratios
tidy(model_full, exponentiate = TRUE) %>%
  kable(
    digits = 3,
    col.names = c("Variable","Odds Ratio","Std Error","z value","p value"),
    caption = "Logistic regression results (full model)"
  )
```

The logistic regression results are summarized in @tbl-glm-full. Film length, release year, and several genres show statistically significant associations with the probability of receiving a high IMDB rating (p \< 0.05).

However, the number of votes is not statistically significant (p = 0.949). Therefore, the variable votes was removed and a simplified model was fitted.

## Predicted Probabilities

To further illustrate the fitted logistic regression model, predicted probabilities were calculated across a range of film lengths while holding release year constant at its mean value.

Predicted probabilities were plotted for several selected genres to visualize how the probability of receiving a high IMDB rating varies with film length and genre.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-pred
#| fig-cap: "Predicted Probability of High IMDb Ratings by Film Length and Genre"
#| fig-align: center
#| fig-width: 6
#| fig-height: 4
#| out-width: 75%

# Fit the simplified logistic regression model (votes removed)
model_final <- glm(
  high_rating ~ length + year + genre,
  data = movies,
  family = binomial
)

# Create new dataset for prediction
newdata <- expand.grid(
  length = seq(0, 200, length.out = 100),
  year = mean(movies$year, na.rm = TRUE),
  genre = c("Comedy", "Documentary", "Drama", "Animation")
)

# Calculate predicted probabilities
newdata$pred_prob <- predict(model_final, newdata = newdata, type = "response")

# Plot predicted probabilities
ggplot(newdata, aes(x = length, y = pred_prob, colour = genre)) +
  geom_line(linewidth = 1.2) +
  labs(
    x = "Film Length (minutes)",
    y = "Predicted Probability",
    colour = "Genre"
  ) +
  theme_bw()
```

@fig-pred shows the predicted probability of receiving a high IMDb rating across different film lengths for selected genres.

Release year is not shown in the figure because it was held constant at its mean value when calculating predicted probabilities. This allows the effects of film length and genre on the predicted probability to be visualized more clearly while still using the simplified logistic regression model.

The predicted probability generally decreases as film length increases, indicating that longer films are less likely to receive ratings above 7. Differences across genres are also evident, with some genres consistently showing higher predicted probabilities than others across the range of film lengths.

## Final Model Interpretation

To aid interpretation of the logistic regression model, the estimated coefficients were transformed into odds ratios by taking their exponential values.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-odds_ratio

# Fit the final logistic regression model
model_final <- glm(
  high_rating ~ length + year + genre,
  data = movies,
  family = binomial
)

# Convert model coefficients to odds ratios for interpretation
tidy(model_final) %>%
  mutate(odds_ratio = exp(estimate)) %>%
  select(term, estimate, odds_ratio, p.value) %>%
  kable(
    digits = 3,
    col.names = c("Predictor", "Estimate", "Odds Ratio", "p value"),
    caption = "Odds ratios from the final logistic regression model"
  )
```

The odds ratios from the final logistic regression model are shown in @tbl-odds_ratio.

Film length shows an odds ratio below 1, indicating that longer films are less likely to receive ratings above 7. In contrast, more recent films and some genres have higher odds of receiving high ratings.

# Model Limitation

Although the logistic regression model provides useful insights, several limitations should be noted.

First, the analysis is based on observational data and therefore cannot establish causal relationships between predictors and IMDb ratings.

Second, some genres contain relatively small numbers of films, which may affect the stability of the estimated coefficients.

Finally, other potentially important factors, such as director reputation, cast popularity or marketing, were not included in the dataset and may also influence film ratings.

# Conclusion

This study examined factors associated with whether a film receives a high IMDB rating.

The results of the logistic regression analysis suggest that film length, release year, and genre are important predictors of high ratings. Longer films tend to have lower probabilities of receiving ratings above 7, while more recently released films show slightly higher probabilities. In addition, some genres such as comedy and documentary are more likely to receive higher ratings compared with the reference genre.

Overall, the findings suggest that both film characteristics and genre differences play a role in determining audience ratings.
