---
title: "Group_8_Analysis"
format:
  html:
    theme: cosmo
    toc: true
    number-sections: true
editor: visual
---

```{r}
#| echo: false
#| warning: false
#| message: false

# Load required packages for data analysis and visualisation
library(ggplot2)   # Data visualisation
library(tidyverse) # Data manipulation and general workflow tools
library(gt)        # Table formatting
library(broom)     # Convert model outputs into tidy data frames
library(knitr)     # Create formatted tables in the report
library(dplyr)     # Data manipulation functions
library(tibble)    # Enhanced data frame tools
```

# Introduction

This study investigates which film characteristics influence whether a film receives an IMDb rating greater than 7. IMDb ratings are widely used as an indicator of audience perception and overall film quality, as they are based on a large number of user evaluations.

Understanding which characteristics are associated with higher ratings may provide insights into what factors contribute to the perceived success of a film. In particular, features such as release year, film length, number of votes and genre may influence how audiences evaluate movies.

To investigate this question, a binary response variable is created to indicate whether a film has an IMDb rating above 7. A logistic regression model within the Generalised Linear Model framework is used to examine how film characteristics influence the probability of receiving a high rating.

# Research Question

The aim of this analysis is to examine which film characteristics are associated with whether a film receives an IMDb rating greater than 7 or not.

To address this question, a binary response variable is considered, taking the value 1 if a film has an IMDb rating above 7 and 0 otherwise. The analysis focuses on whether release year, film length, production budget, number of votes, and genre are related to the probability of receiving a high rating.

Given that the response variable is binary, the analysis employs logistic regression as part of the Generalised Linear Model framework.

# The Dataset {#sec-dataset}

The dataset used in this analysis comes from the IMDb film database, which provides a large collection of information about films and their characteristics. The data analysed here represent a subset of this database containing key variables relevant to film popularity and ratings.

## Inspect the dataset {#sec-insp}

```{r}
#| echo: false
#| warning: false
#| message: false

# Read the movie dataset from the CSV file
movies <- read.csv("dataset08.csv")
```

After loading the dataset, the first few observations are displayed to provide an initial overview of the data.

```{r}
#| echo: false
#| warning: false
#| message: false

# Display the first six rows to inspect the structure of the dataset
head(movies)
```

The structure of the dataset is examined to identify the variables and their data types.

```{r}
#| echo: false
#| warning: false
#| message: false
#| results: false

# Examine the structure of the dataset
# This shows variable types and basic information about each column
str(movies)
```

Based on the structure output, the dataset contains 2,847 observations and seven variables. Specifically, film_id, year, length, and votes are stored as integer variables, while budget and rating are numeric variables. The genre variable is recorded as a character variable representing film categories. The structure output also suggests that some variables, such as length, contain missing values (NA).

To further explore the distribution of the variables in the dataset, summary statistics are computed.

```{r}
#| echo: false
#| warning: false
#| message: false
#| results: false

# Produce basic summary statistics for all variables
summary(movies)
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-numvars
#| tbl-cap: "Summary statistics for numerical variables"

# Select numerical variables for descriptive statistics
numeric_vars <- movies %>% 
  select(film_id, year, length, budget, votes, rating)

# Compute key descriptive statistics for each variable
numeric_stats <- rbind(
  summarise(numeric_vars, across(everything(), ~min(.,            na.rm = TRUE))), # Minimum
  summarise(numeric_vars, across(everything(), ~quantile(., 0.25, na.rm = TRUE))), # 1st quartile
  summarise(numeric_vars, across(everything(), ~median(.,         na.rm = TRUE))), # Median
  summarise(numeric_vars, across(everything(), ~mean(.,           na.rm = TRUE))), # Mean
  summarise(numeric_vars, across(everything(), ~quantile(., 0.75, na.rm = TRUE))), # 3rd quartile
  summarise(numeric_vars, across(everything(), ~max(.,            na.rm = TRUE))), # Maximum
  summarise(numeric_vars, across(everything(), ~sum(is.na(.))))                     # Number of missing values
) 

# Transpose the table to present variables as rows
numeric_stats <- numeric_stats %>%
  t() %>%
  as.data.frame() %>%
  setNames(c("Min", "1st Qu", "Median", "Mean", "3rd Qu", "Max", "Missing")) %>%
  rownames_to_column("Variable") %>%
  mutate(across(-Variable, ~round(as.numeric(.), 2))) # Round results for readability

# Display the formatted table
kable(numeric_stats)
```

@tbl-numvars presents the summary statistics for the numerical variables included in the analysis.

As shown in @tbl-numvars, the release year of the films spans from 1898 to 2005, suggesting that the dataset covers a long historical period. Film length ranges from 1 to 480 minutes. The production budget ranges between 2.5 and 22.3, while the number of votes shows particularly large variation, with a maximum value of 149,494. In contrast, film ratings appear to be relatively concentrated, with values ranging from 0.8 to 9.2. It is also noteworthy that the variable *length* contains 131 missing observations, whereas the remaining numerical variables do not exhibit missing values.

In addition to the numerical variables, the distribution of film genres is also examined.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-genre
#| tbl-cap: "Distribution of films across genres"

# Count the number of films in each genre
genre_summary <- movies %>%
  count(genre, sort = TRUE)

# Display the genre frequency table
kable(genre_summary, caption = "Number of films by genre")
```

@tbl-genre reports the number of films in each genre.

@tbl-genre presents the distribution of films across seven genres in the dataset. Action, Drama, and Comedy account for the largest number of observations, whereas genres such as Animation, Documentary, Short, and Romance appear less frequently. This suggests that the dataset is dominated by a few major genres.

## Variable descriptions

In this dataset, each film is represented by several variables describing its main characteristics. @tbl-variable shows the main variables used, including their data types and brief descriptions.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-variable
#| tbl-cap: "Description of variables in the dataset"


# Create a data frame that summarises the main variables in the dataset
var_desc <- data.frame(
  Variable = c("film_id", "year", "length", "budget", "votes", "genre", "rating"),
  Type = c("Integer", "Integer", "Numeric", "Numeric", "Integer", "Character", "Numeric"),
  Description = c(
    "The unique identifier for the film",
    "Year of release of the film in cinemas",
    "Duration (in minutes)",
    "Budget for the films production (in $1000000s)",
    "Number of positive votes received by viewers",
    "Genre of the film",
    "IMDb rating from 0-10"
  )
)


# Display the variable description table
kable(var_desc)
```

@tbl-variable shows the variables included in the dataset.

The dataset contains both numerical and categorical variables describing different aspects of each film. These variables will later be used as predictors in the statistical model.

# Exploratory Analysis

Before fitting the GLM, exploratory analysis was carried out to examine the relationships between the response variable and the candidate predictors. This helps to identify broad patterns in the data and provides preliminary support for variable selection in the modelling stage.

```{r}
#| echo: false
#| warning: false
#| message: false

# Create a binary response variable for the logistic regression
# 1 = IMDb rating above 7
# 0 = IMDb rating 7 or below
movies$high_rating <- ifelse(movies$rating > 7, 1, 0)
```

## Release Year and High IMDb Ratings

Release year may influence how films are rated. To explore this, films were grouped by release decade and the proportion of highly rated films (IMDb rating \> 7) was calculated for each group.

Very early decades contained only a small number of films, so films released before 1910 were excluded from this plot to provide a more stable comparison across release periods.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-explo-year
#| fig-cap: "Proportion of Highly Rated Films by Release Decade"
#| fig-align: center
#| fig-width: 6
#| fig-height: 4
#| out-width: 75%

# Create a dataset summarising high ratings by release decade
movies_year <- movies %>%
  filter(year >= 1910) %>%                     # Remove very early films with few observations
  mutate(decade = floor(year/10)*10) %>%       # Convert year into decade groups
  group_by(decade) %>%
  summarise(
    proportion_high = mean(high_rating, na.rm = TRUE), # Proportion of high ratings
    count = n()                                         # Number of films in each decade
  )

# Plot the proportion of high-rated films across decades
ggplot(movies_year, aes(x = decade, y = proportion_high)) +
  geom_line(colour = "mediumpurple", linewidth = 1.2) +
  geom_point(colour = "gold2", size = 2.5) +
  geom_hline(yintercept = mean(movies$high_rating),     # Overall proportion reference line
             linetype = "dashed",
             colour = "coral",
             linewidth = 0.8) +
  labs(
    x = "Decade",
    y = "Proportion of Films with Rating > 7"
  ) +
  theme_bw()
```

@fig-explo-year shows the proportion of highly rated films across release decades. The dashed horizontal line represents the overall proportion of films with ratings above 7 in the dataset.

The proportion of highly rated films appears to vary across decades. Earlier decades tend to show somewhat higher proportions, while films released between the 1950s and 1980s appear to have relatively lower proportions. However, some early decades contain relatively few films, so these proportions should be interpreted with caution.

## Film Length and IMDb Ratings

Film length may influence audience perception of a movie. To explore this relationship, film length is treated as a continuous variable and the relationship between film length and the probability of receiving a high rating is visualised using a logistic smoothing curve.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-explo-length
#| fig-cap: "Relationship Between Film Length and the Probability of High IMDb Ratings"
#| fig-align: center
#| fig-width: 6
#| fig-height: 4
#| out-width: 75%

# Remove observations with missing film length values
movies_length <- subset(movies, !is.na(length))

# Plot relationship between film length and probability of high rating
ggplot(movies_length, aes(x = length, y = high_rating)) +
  geom_jitter(height = 0.05, alpha = 0.3, colour = "pink") + # Scatter points with jitter
  geom_smooth(method = "glm",                                # Logistic regression smoothing
              method.args = list(family = "binomial"),
              colour = "navy") +
  labs(
    x = "Film Length (minutes)",
    y = "Probability of High Rating"
  ) +
  theme_bw()
```

@fig-explo-length shows the relationship between film length and the probability of receiving a high IMDb rating.

The probability of receiving a rating above 7 appears to be higher for shorter films and gradually decreases as film length increases. This suggests that film length may be associated with high ratings, although the pattern may partly reflect differences in genre composition across films of different lengths.

## Production budget and IMDb Ratings

Film budget may also influence how audiences perceive a movie. To explore this possibility, the distribution of film budgets and their relationship with IMDb ratings were examined.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-explo-budget
#| fig-cap: Relationship between film budget and the probability of receiving a high IMDb rating
#| fig-align: center
#| fig-width: 6
#| fig-height: 4
#| out-width: 75%

# Plot relationship between film budget and probability of high rating
ggplot(movies,
       aes(x = budget,
           y = high_rating)) +
  geom_jitter(height = 0.05, alpha = 0.3) +                  # Scatter points
  geom_smooth(
    method = "glm",                                          # Logistic smoothing curve
    method.args = list(family="binomial"),
    colour = "red"
  ) +
  theme_bw()
```

@fig-explo-budget shows the relationship between film production budget and the probability of receiving a high IMDb rating. Each point represents an individual film, and the red curve corresponds to the fitted logistic regression line.

The fitted curve shows a gradual upward trend, suggesting that films with larger budgets may be somewhat more likely to receive high ratings. This suggests that budget may be positively associated with the likelihood of a film receiving a high rating.

## Votes and High IMDb Ratings

The number of votes may reflect a film’s popularity or level of audience engagement. Since the distribution of votes is highly skewed, with some films receiving far more votes than others, direct comparisons using raw vote counts are difficult. Therefore, films were divided into five groups based on vote quantiles so that each group contains approximately the same number of observations. This grouping allows the proportion of highly rated films to be compared across different levels of audience engagement while reducing the influence of extreme vote counts.

The approximate vote ranges for the five groups are:\
Very Low: 5-10 votes\
Low: 10-20 votes\
Medium: 20-47 votes\
High: 47-174 votes\
Very High: above 174 votes.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-explo-votes
#| fig-cap: Proportion of High IMDb Ratings Across Vote Groups
#| fig-align: center
#| fig-width: 6
#| fig-height: 4
#| out-width: 75%

# Divide vote counts into five groups using quantiles
vote_breaks <- quantile(movies$votes, probs = seq(0,1,0.2), na.rm = TRUE)

# Create a categorical variable representing vote groups
movies$vote_group <- cut(
  movies$votes,
  breaks = vote_breaks,
  include.lowest = TRUE,
  labels = c("Very Low", "Low", "Medium", "High", "Very High")
)

# Calculate proportion of high ratings in each vote group
prop_table <- aggregate(high_rating ~ vote_group,
                        data = movies,
                        FUN = mean)

# Plot proportion of high ratings by vote group
ggplot(prop_table, aes(x = vote_group, y = high_rating)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_hline(yintercept = mean(movies$high_rating), # Overall proportion
             colour = "red",
             linetype = "dashed",
             linewidth = 1) +
  labs(
    x = "Vote Groups",
    y = "Proportion of Films with Rating > 7"
  ) +
  theme_minimal()
```

@fig-explo-votes shows the proportion of highly rated films across vote groups. The dashed horizontal line represents the overall proportion of films with ratings above 7 in the dataset.

The proportion of highly rated films appears to vary across vote groups, with lower-vote groups showing slightly higher proportions than higher-vote groups. Although the pattern is not especially strong, it suggests that the relationship between vote count and high ratings may not be straightforward.

## Film Genre and High IMDb Ratings

Film genre may influence how movies are perceived by audiences. Different genres may attract different audiences and may vary in storytelling style, which could affect IMDb ratings.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-explo-genre
#| fig-cap: "Proportion of Highly Rated Films by Genre"
#| fig-align: center
#| fig-width: 6
#| fig-height: 4
#| out-width: 75%

# Calculate proportion of highly rated films for each genre
genre_prop <- movies %>%
  group_by(genre) %>%
  summarise(
    proportion_high = mean(high_rating, na.rm = TRUE),
    count = n()
  )

# Plot high rating proportion by genre
ggplot(genre_prop,
       aes(x = reorder(genre, proportion_high),
           y = proportion_high)) +
  geom_col(fill = "seagreen") +
  geom_hline(yintercept = mean(movies$high_rating, na.rm = TRUE), # Overall proportion
             colour = "darkred",
             linetype = "dashed",
             linewidth = 1) +
  coord_flip() +
  labs(
    x = "Genre",
    y = "Proportion of Films with Rating > 7"
  ) +
  theme_bw()
```

@fig-explo-genre shows substantial variation in the proportion of highly rated films across genres. Some genres have much higher proportions of films with ratings above 7 than others.

This suggests that genre may be an important predictor of high ratings. However, some genres contain relatively few observations, so the corresponding proportions may be less stable and should be interpreted cautiously.

# Model Fitting

## Data Preprocessing

Before fitting the logistic regression models, a modelling dataset was created by selecting the variables required for the analysis and removing observations with missing values. In this analysis, the response variable is high_rating, while the candidate predictors are release year, film length, production budget, number of votes, and genre.

```{r}
#| echo: false
#| warning: false
#| message: false

# Create modelling dataset by selecting relevant variables
movies_model <- movies %>%
  select(high_rating, year, length, budget, votes, genre) %>%
  na.omit()   # Remove rows with any missing values

# Convert genre into a categorical variable
movies_model$genre <- factor(movies_model$genre)

```

The raw distribution of film genres was previously presented in @tbl-genre, and the proportion of highly rated films across genres was illustrated in @fig-explo-genre. However, the logistic regression model uses the complete dataset after removing rows with missing values. Therefore, the reference category for genre should be chosen based on the final modelling sample rather than the raw dataset.

To assess which genre is most suitable as the reference category, the distribution of the binary outcome variable (high_rating) across genres is examined in the modelling dataset.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-genre-model
#| tbl-cap: "Distribution of high ratings across genres in the modelling dataset"


# Summarise the outcome distribution for each genre in the modelling dataset
genre_outcome_model <- movies_model %>%
  group_by(genre) %>%
  summarise(
    n = n(),  # n = total number of films in the genre  
    high_rating_0 = sum(high_rating == 0),  # high_rating_0 = number of films with rating <= 7
    high_rating_1 = sum(high_rating == 1),  # high_rating_1 = number of films with rating > 7
    prop_high = mean(high_rating)  # prop_high = proportion of films with rating > 7
  ) %>%
  arrange(desc(n))


# Display the summary table
kable(genre_outcome_model, digits = 3)
```

@tbl-genre-model shows that the genre categories differ significantly in both sample size and outcome distribution. Action has the largest number of observations in the modelling dataset, which makes it a stable category for comparison. In addition, Action contains both low-rated and high-rated films, rather than being concentrated almost entirely in one outcome group.

Some genres exhibit much more extreme outcome distributions. For example, Romance contains no films with ratings above 7, while Short and Documentary are heavily concentrated in the high-rating group. Among the remaining genres, Drama is strongly skewed towards low-rated films, whereas Comedy and Animation are relatively more concentrated in higher ratings. Compared with these categories, Action represents a more moderate and less extreme distribution.

To visualise these differences more clearly, @fig-genre-model shows the proportion of low-rated and high-rated films within each genre.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-genre-model
#| fig-cap: "Proportion of low-rated and high-rated films within each genre in the modelling dataset"
#| fig-align: center
#| fig-width: 6
#| fig-height: 4
#| out-width: 75%

# Create a proportional bar chart to show the composition of high_rating within each genre
ggplot(movies_model, 
       aes(x = reorder(genre, -table(genre)[genre]), fill = factor(high_rating))) +
  geom_bar(position = "fill") +   # Show proportions instead of counts
  labs(
    x = "Genre",
    y = "Proportion",
    fill = "High Rating"
  ) +
  scale_fill_discrete(
    labels = c("0 = Rating <= 7", "1 = Rating > 7")
  ) +
  theme_bw()
```

For these reasons, Action was selected as the reference category for genre. This choice is supported by three aspects: it remains the largest genre in the modelling dataset, it provides a relatively stable baseline, and it allows for meaningful comparisons across genres.

```{r}
#| echo: false
#| warning: false
#| message: false

# Set Action as the reference category for genre
movies_model$genre <- relevel(movies_model$genre, ref = "Action")

```

After preprocessing, the modelling dataset is ready for logistic regression analysis, with genre coded as a categorical variable and Action defined as the baseline category for comparison.

## Fitting Full Model

A full logistic regression model was first fitted including the number of votes, film length, release year, production budget, and genre as predictors of whether a film received an IMDb rating above 7.

The full model is specified as:

$$
\text{high\_rating} \sim \text{year} + \text{length} + \text{budget} + \text{votes} + \text{genre}
$$

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-full-model

# Fit the full logistic regression model
model_full <- glm(
  high_rating ~ year + length + budget + votes + genre,
  data = movies_model,
  family = binomial
)


# Display model results in a formatted table
tidy(model_full) %>%
  kable(
    digits = 3,
    col.names = c("Variable", "Estimate", "Std Error", "z value", "p value"),
    caption = "Logistic regression results (full model)"
  )

```

The logistic regression results are summarised in @tbl-full-model. The number of votes is not statistically significant, while film length, release year, production budget, and several genre categories show significant associations with the probability of receiving a high IMDb rating.

## Fitting Reduced Model

Since the number of votes was not statistically significant in the full model, it was removed and a reduced logistic regression model was fitted using the remaining predictors: film length, release year, production budget, and genre.

The reduced model is specified as:

$$
\text{high\_rating} \sim \text{year} + \text{length} + \text{budget} + \text{genre}
$$

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-reduced-model

# Fit reduced model excluding votes
model_reduced <- glm(
  high_rating ~ year + length + budget + genre,
  data = movies_model,
  family = binomial
)

# Display reduced model results
tidy(model_reduced) %>%
  kable(
    digits = 3,
    col.names = c("Variable", "Estimate", "Std Error", "z value", "p value"),
    caption = "Logistic regression results (reduced model)"
  )
```

The reduced model results are shown in @tbl-reduced-model. Film length, release year, production budget, and several genre categories remain statistically significant predictors of whether a film achieves an IMDb rating above 7.

## Model comparison

### Coefficient Comparison Table

To compare the full and reduced models, the estimated coefficients are presented side by side in @tbl-model-comp.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-model-comp

# Extract coefficients from full model
full_coef <- tidy(model_full) %>%
  select(term, estimate) %>%
  rename(Full_Model = estimate)

# Extract coefficients from reduced model
reduced_coef <- tidy(model_reduced) %>%
  select(term, estimate) %>%
  rename(Reduced_Model = estimate)

# Combine results into a comparison table
comparison <- full_coef %>%
  full_join(reduced_coef, by = "term")

# Display comparison table
kable(
  comparison,
  digits = 3,
  col.names = c("Predictor", "Full Model", "Reduced Model"),
  caption = "Comparison of coefficients between full and reduced models"
)
```

As shown in @tbl-model-comp, removing the non-significant variable votes leads to only negligible changes in the estimated coefficients of the remaining predictors. In particular, the coefficients for release year, film length, production budget, and the genre categories remain very similar in magnitude and direction across the two models. This indicates that excluding votes has little impact on the estimated relationships between the other predictors and the probability of receiving a high IMDB rating.

### Model Fitting Statistics Table

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-model-statis


# Create a summary table of model fit statistics
# AIC is used to compare model fit while accounting for model complexity
# Residual deviance and residual degrees of freedom are also reported
model_fit <- tibble(
  Model = c("Full model", "Reduced model"),
  AIC = c(AIC(model_full), AIC(model_reduced)),
  Residual_Deviance = c(deviance(model_full), deviance(model_reduced)),
  DF = c(df.residual(model_full), df.residual(model_reduced))
)


# Display the model fit statistics table
kable(
  model_fit,
  digits = 3,
  caption = "Model fit statistics for the full and reduced models"
)
```

The model fitting statistics for the full and reduced models are presented in @tbl-model-statis. The reduced model has a slightly lower AIC than the full model, which indicates a better balance between model fit and model complexity. Although the residual deviance of the reduced model is marginally higher than that of the full model, the difference is negligible. This indicates that removing the variable votes leads to almost no loss of explanatory power while producing a simpler model.

## Final Model Selection

In the full model, the coefficient for the number of votes was not statistically significant. After removing this variable, the reduced model produced a slightly lower AIC and almost the same residual deviance as the full model. In addition, the comparison of coefficients between the two models showed that the estimated effects of the remaining predictors changed very little after excluding votes. Therefore, the reduced model was selected as the final model because it achieves a slight improvement in overall model performance while maintaining a very similar level of fit, and provides a more parsimonious representation of the data.

The final selected model is:

$$
\text{high\_rating} \sim \text{year} + \text{length} + \text{budget} + \text{genre}
$$

## Odds Ratios of the Final Model

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-or-reduced

# Convert logistic regression coefficients to odds ratios
or_reduced <- tidy(model_reduced, conf.int = TRUE) %>%
  mutate(
    OR = exp(estimate),       # Odds ratio
    CI_low = exp(conf.low),   # Lower confidence interval
    CI_high = exp(conf.high)  # Upper confidence interval
  ) %>%
  select(term, OR, std.error, statistic, p.value, CI_low, CI_high)

# Display odds ratio table
kable(
  or_reduced,
  digits = 3,
  col.names = c("Predictor", "OR", "SE", "z", "p", "CI low", "CI high"),
  caption = "Odds ratios for the reduced logistic regression model"
)
```

@tbl-or-reduced presents the odds ratios from the final logistic regression model. The odds ratios are obtained by exponentiating the estimated coefficients and provide an interpretable measure of how each predictor affects the odds that a film receives an IMDb rating above 7, holding the other variables constant.

## Predicted Probabilities

To further illustrate the final logistic regression model, predicted probabilities were calculated using the reduced model. Film length was allowed to vary across its observed range, while release year and production budget were held constant at their mean values.

Predicted probabilities were generated for several selected genres in order to visualise how the probability of receiving an IMDB rating above 7 changes with film length and genre.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-pred-prob
#| fig-cap: "Predicted Probability of High IMDB Rating by Film Length and Genre"
#| fig-align: center
#| fig-width: 6
#| fig-height: 4
#| out-width: 75%

# Create new dataset for prediction
newdata <- expand.grid(
  length = seq(
    min(movies_model$length, na.rm = TRUE),
    max(movies_model$length, na.rm = TRUE),
    length.out = 100
  ),
  year = mean(movies_model$year, na.rm = TRUE),   # Fix year at mean value
  budget = mean(movies_model$budget, na.rm = TRUE), # Fix budget at mean value
  genre = c("Action", "Comedy", "Documentary", "Drama", "Short") # Selected genres
)

# Generate predicted probabilities from the reduced model
newdata$pred_prob <- predict(
  model_reduced,
  newdata = newdata,
  type = "response"
)

# Plot predicted probabilities
ggplot(newdata, aes(x = length, y = pred_prob, colour = genre)) +
  geom_line(linewidth = 1.2) +
  labs(
    x = "Film Length (minutes)",
    y = "Predicted Probability",
    colour = "Genre"
  ) +
  theme_bw()
```

@fig-pred-prob shows the predicted probability that a film receives an IMDB rating above 7 across different film lengths for several selected genres.

The predicted probabilities decrease as film length increases, indicating that longer films are generally less likely to receive ratings above 7 according to the fitted model. This pattern is consistent with the negative coefficient of film length observed in the logistic regression results.

Differences between genres can be clearly seen. Documentary films consistently have the highest probability of receiving a rating above 7 across most film lengths, with Short films also showing relatively high probabilities. Comedy films show moderate probabilities, followed by Action films. In contrast, Drama films have the lowest probability of receiving a rating above 7 across the range of film lengths considered.

Overall, the figure illustrates how the final logistic regression model combines the effects of film length and genre to predict the probability that a film achieves an IMDB rating above 7, while holding the other predictors constant.

# Model Limitation

This analysis has several limitations. First, the dataset contains only a limited set of film characteristics, including release year, film length, production budget, number of votes, and genre. Other potentially important factors, such as director, cast, country of production, awards, or marketing exposure, are not included in the dataset. As a result, the model may omit variables that are also related to whether a film receives a high IMDb rating.

Second, missing values in film length were handled by removing incomplete observations. Although this provides a simple and consistent modelling dataset, it reduces the sample size and may introduce bias if the missing values are not completely random.

Third, some genres contain relatively few films compared with the larger categories. This may lead to less stable coefficient estimates for those smaller groups, so the corresponding genre effects should be interpreted with caution.

# Conclusion

This analysis examined which film characteristics are associated with whether a film receives an IMDb rating greater than 7. Because the response variable was binary, logistic regression was used to model the probability of a film achieving a high rating.

A full model including number of votes, film length, release year, production budget, and genre was first fitted. Since the number of votes was not statistically significant, a reduced model excluding this variable was selected as the final model. The reduced model also showed a slightly lower AIC, indicating a better balance between model fit and model simplicity.

Overall, the results suggest that film length, release year, production budget, and genre are important predictors of whether a film receives an IMDb rating above 7. In particular, longer films were associated with a lower predicted probability of achieving a high rating, while higher production budgets and more recent release years were associated with a greater likelihood of receiving a rating above 7. Clear differences across genres were also observed, with some genres showing consistently higher predicted probabilities than others.

In summary, the final logistic regression model provides a useful statistical framework for understanding how several film characteristics are associated with the probability of receiving a high IMDb rating.
